---
title: "Revolving MSMR with MLHO"
author: "Jonas Hügel"
date: "2024-03-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Load packages

```{r package loading}

##update/install MLHO
devtools::install_github("clai-group/mlho")

# load MLHO, afterwards source the MSMR.lite.R file to overwrite the MSMR.lite function
# in the package with the updated one (the encounter functionality is only available in the R file)
library(mlho)

#load and install required dependecies
pacman::p_load(data.table, devtools, backports, Hmisc, tidyr,dplyr,ggplot2,plyr,scales,readr,
                httr, DT, lubridate, tidyverse,reshape2,foreach,doParallel,caret,gbm,lubridate,praznik)
               
```

## Prepare the data

```{r data_loading}

dbmart <- mlho::incident_dbmart
labelDT <- mlho::incident_labeldt
#TODO sex_colum contains only 1 value, resuling in warnings during training -> change in incident datafile from the package 
dems <- mlho::incident_dems
```

##  determine the split for test and train using a 70-30 ratio

```{r train_test_split}
uniqpats <- c(as.character(unique(dbmart$patient_num)))


test_ind <- sample(uniqpats,
                   round(.3*length(uniqpats)))

```

## transform train data
```{r transform_train_data}

dat.train  <- subset(dbmart,!(dbmart$patient_num %in% c(test_ind)))
data.table::setDT(dat.train)
#values must be in column named value
dat.train[,value := 1]
#dat.train$value.var <- 1
uniqpats.train <- c(as.character(unique(dat.train$patient_num)))

MLHO.dat <- dat.train
patients <- uniqpats.train
sparsity=0.05
jmi=TRUE
topn=200
patients <- uniqpats.train
multicore=FALSE
encounterLevel=TRUE
valuesToMerge = TRUE
dat.train <- mlho::MSMR.lite(dat.train,
                   labelDT,
                   binarize=FALSE,
                   sparsity=0.05,
                   jmi=TRUE,
                   topn=200,
                   patients <- uniqpats.train,
                   multicore=FALSE,
                   encounterLevel=TRUE,
                   valuesToMerge = TRUE)

```

## transform test data
```{r transforma_test_data} 

dat.test <- subset(dbmart,dbmart$patient_num %in% c(test_ind))
uniqpats.test <- c(as.character(unique(dat.test$patient_num)))
# remove phenx not required to create the encounter based phenx 
# (remove _last, _past and _history from the colnames to determine the phenxs)
dat.train.colnames <- vapply(strsplit(colnames(dat.train),"_"),`[`, 1, FUN.VALUE=character(1))
dat.test <- subset(dat.test,dat.test$phenx %in% dat.train.colnames)
setDT(dat.test)
#values must be in column named value
dat.test$value <- 1

# important to have a value and phenx column to merge
dat.test <- MSMR.lite(MLHO.dat=dat.test,
                       patients = uniqpats.test,
                       sparsity=NA,
                       jmi = FALSE,
                       labels = labelDT,
                       encounterLevel = TRUE,
                       valuesToMerge = TRUE)

# remove sparse and not relevant _past, _last _history phenx according to the train data
dat.test <- dat.test %>% select(one_of(colnames(dat.train)))

```

## update dems and labelDT
``` {r update dems and labels}

dems <- dems %>%
  merge(labelDT,by = "patient_num") %>%
  mutate(patient_num = paste0(patient_num,"_" ,start_date)) %>%
  select(-start_date, -label)

# merge patientnum)and encounter date in labelDT
labeDT <- labelDT %>%
  mutate(patient_num = paste0(patient_num,"_" ,start_date))  %>%
  select(-start_date)

```

## train model
we may want to reduce the output of this cell
```{r training}


model.test <- mlearn(dat.train,
                     dat.test,
                     dems=dems,
                     save.model=FALSE,
                     classifier="gbm",
                     note="mlho_terst_run",
                     cv="cv",
                     nfold=5,
                     aoi="random phenx from dbmart",
                     multicore=FALSE)

```




## visualize results
Here we create a plot of the feature importance scores for each of the top (here we have `r nrow(model.test$coefficients)`) predictors identified by MLHO.

To do so, let's map the concept codes to their "English" translation. *That's why we kept that 4th column called `description` in `dbmart`.*


```{r visualisazation}
#TODO FIX! copie from original vignette

model.test$features

#TODO fix merge, cut past/last/history before merging the features with the concepts and append afterwards
dbmart.concepts <- dbmart[!duplicated(paste0(dbmart$phenx)), c("phenx","DESCRIPTION")]
mlho.features <- data.frame(merge(model.test$features,dbmart.concepts,by.x="features",by.y = "phenx"))
datatable(dplyr::select(mlho.features,features,DESCRIPTION,`Feature importance`=Overall), options = list(pageLength = 5), filter = 'bottom')
```

now visualizing feature importance
```{r visualisazation_fig,fig.align='center',fig.width=7,fig.height=7}
#TODO FIX
(plot<- ggplot(mlho.features) +
    geom_segment(
      aes(y = 0,
          x = reorder(DESCRIPTION,Overall),
          yend = Overall,
          xend = DESCRIPTION),
      size=0.5,alpha=0.5) +
    geom_point(
      aes(x=reorder(DESCRIPTION,Overall),y=Overall),
      alpha=0.5,size=2,color="red") +
    theme_minimal()+
   coord_flip()+
    labs(y="Feature importance",x=""))
```

