---
title: "Benchmark"
format: html
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages

```{r,echo=TRUE, results = 'hide'}

##update/install MLHO
devtools::install_github("clai-group/mlho")

# load MLHO, afterwards source the MSMR.lite.R file to overwrite the MSMR.lite function
# in the package with the updated one (the encounter functionality is only available in the R file)
library(mlho)

#load and install required dependecies
pacman::p_load(data.table, devtools, backports, Hmisc, tidyr,dplyr,ggplot2,plyr,scales,readr, httr, DT, lubridate, DALEX, tidyverse,reshape2,foreach,doParallel,caret,gbm,lubridate,praznik)
library(counterfactuals)
library(iml)
```

## Prepare the data

We load several datasets from the MLHO package, including incident data and demographic information.

`dbmart` consists of patient ID (patient_num) and associated phenotypes (phenx). Each patient can have multiple features, including different diagnostic events or conditions.

`labelDT` includes patient ID (patient_num), the start date of each event (start_date), and a binary label (label) indicating the outcome of interest.

`dems` contains dempgraphic information for each patient.

```{r data_loading}
dbmart <- mlho::incident_dbmart
labelDT <- mlho::incident_labeldt
labelDT <- labelDT %>% mutate(o_date=case_when(label == 0 ~ start_date, label == 1 ~ start_date+sample(1:40,1)))
labelDT$start_date= as.Date(labelDT$start_date)
dems <- mlho::incident_dems
```

```{r}
head(dbmart)
head(labelDT)
```

## Splitting data into training and testing sets using a 70-30 ratio

We extract a unique list of "patient_num" from `dbmart.` Using the list of unique patient ID, we randomly select 30% of these patients to include in our test set.

```{r train_test_split}
uniqpats <- c(as.character(unique(dbmart$patient_num)))


test_ind <- sample(uniqpats,
                   round(.3*length(uniqpats)))

```

## Transform train data

After splitting the data into training and testing sets, the next step is to transform the data to ensure that the data aligns with the requirements of the modeling functions in the MLHO package.

```{r transform_train_data}
dat.train  <- subset(dbmart,!(dbmart$patient_num %in% c(test_ind)))
data.table::setDT(dat.train)
#values must be in column named value
dat.train[,value := 1]
uniqpats.train <- c(as.character(unique(dat.train$patient_num)))
```

The `benchmark.revolving` function in the `MLHO` package generates a baseline, atemporal representation of patient data for machine learning. Unlike the TLDR algorithm, which labels each clinical record with temporal context (e.g., "history," "past," and "last"), this benchmarking approach ignores the timing of events. Instead, it simply counts the total number of times each clinical feature appears across all of a patient's encounters. For each patient, the output is a wide-format table where each column represents a unique feature, and each cell contains the total count of that feature---aggregated over time, without regard to when it occurred.

```{r}
MLHO.dat <- dat.train
labels = labelDT
patients <- uniqpats.train
binarize=T
sparsity=0.05 ## Sample size * sparsity. Don't pick a too small value to avoid overfitting
jmi=TRUE
topn=50
patients <- uniqpats.train
multicore=T
valuesToMerge = F
timeBufffer=c(h=0,p=0,l=0,o=-30)
dat.train <- benchmark.revolving(MLHO.dat,
                                 labels,
                                 binarize,
                                 sparsity,
                                 jmi,
                                 topn,
                                 patients <- uniqpats.train,
                                 multicore=FALSE,
                                 valuesToMerge = TRUE,
                                 timeBufffer)
rank = dat.train$rank
dat.train = dat.train$AVR
```

## Transform test data

We repeat the data processing and transformation again on the test set.

```{r}
dat.test <- subset(dbmart,dbmart$patient_num %in% c(test_ind))
uniqpats.test <- c(as.character(unique(dat.test$patient_num)))
# remove phenx not required to create the encounter based phenx 
# (remove _last, _past and _history from the colnames to determine the phenxs)
dat.train.colnames <- vapply(strsplit(colnames(dat.train),"_"),`[`, 1, FUN.VALUE=character(1))
dat.test <- subset(dat.test,dat.test$phenx %in% dat.train.colnames)
setDT(dat.test)
#values must be in column named value
dat.test$value <- 1

MLHO.dat.test = dat.test

# important to have a value and phenx column to merge
dat.test <- benchmark.revolving(MLHO.dat=dat.test,
                                patients = uniqpats.test,
                                sparsity=NA,
                                jmi = FALSE,
                                labels = labelDT,
                                valuesToMerge = TRUE,
                                binarize = F,
                                timeBufffer)

# remove sparse and not relevant _past, _last _history phenx according to the train data
dat.test <- dat.test %>% select(one_of(colnames(dat.train)))
```

## Update demographics and labels Data

The `dems` dataset, which contains demographic information, is updated to include relevant labels from `labelDT`. This integration involves merging both datasets by "patient_num", then modifying the "patient_num" to include the "start_date" for a unique identifier per patient encounter.

```{r}
dems <- dems %>%
  merge(labelDT,by = "patient_num") %>%
  mutate(patient_num = paste0(patient_num,"_" ,start_date)) %>%
  select(-start_date, -label)
```

Similarly, `labelDT` is updated to concatenate "patient_num" with "start_date" to create a unique identifier for each patient's encounter, which simplifies subsequent merging and data handling processes. The "start_date" column is then removed to clean up the dataset:

```{r}
# merge patientnum and encounter date in labelDT
labelDT <- labelDT %>%
  mutate(patient_num = paste0(patient_num,"_" ,start_date))  %>%
  select(-start_date)

```

## Train model

We use the `mlearn` function to do the modeling, which includes training the model and testing it on the test set.

```{r}
## we may want to reduce the output of this cell
model.test <- mlearn(dat.train,
                     dat.test,
                     dems=NULL,
                     save.model=FALSE,
                     classifier="gbm",
                     note="mlho_test_run",
                     cv="cv",
                     nfold=5,
                     aoi="random phenx from dbmart",
                     multicore=FALSE,
                     calSHAP = T,
                     counterfactual = T,
                     save.model.counterfactual = F)

```

## Visualize results

Here we create a plot of the feature importance scores for each of the top (here we have `r nrow(model.test$coefficients)`) predictors identified by MLHO.

To do so, let's map the concept codes to their "English" translation. *That's why we kept that 4th column called `description` in `dbmart`.*

```{r}
features = model.test$features
features$features.new <- sub("_.*", "", features$features )
dbmart.concepts <- dbmart[!duplicated(paste0(dbmart$phenx)), c("phenx","DESCRIPTION")]
mlho.features <- data.frame(merge(features,dbmart.concepts,by.x="features.new",by.y = "phenx"))
datatable(dplyr::select(mlho.features,features,DESCRIPTION,`Feature importance`=Overall), options = list(pageLength = 5), filter = 'bottom')
```

now visualizing feature importance

```{r visualisazation_fig,fig.align='center',fig.width=11,fig.height=5}
(plot<- ggplot(mlho.features) +
    geom_segment(
      aes(y = 0,
          x = reorder(DESCRIPTION,Overall),
          yend = Overall,
          xend = DESCRIPTION),
      size=0.5,alpha=0.5) +
    geom_point(
      aes(x=reorder(DESCRIPTION,Overall),y=Overall),
      alpha=0.5,size=2,color="red") +
    theme_minimal()+
   coord_flip()+
    labs(y="Feature importance",x=""))
```

## SHAP value and visualization

When setting `calSHAP=TRUE`, SHAP values are also calculated to explain the output of models as shown below.

```{r}
shap_value <- model.test$shap
head(as.data.frame(shap_value))
```

```{r,fig.width=11,fig.height=5}
plot(shap_value)
dbmart.concepts.new = mlho.features %>% select(features, DESCRIPTION)
colnames(dbmart.concepts.new) = c("phenx", "DESCRIPTION")
mshapviz(shap_value, dbmart.concepts.new, plot_type = "waterfall", top_n = 6, num = 1)
mshapviz_all(shap_value, dbmart.concepts.new, top_n = 5)
```
